### 2017-01-28 paper/rebuttal.txt

- First, who are the intended users of the proposed tool? Are they end users
  who might not have technical knowledge about data transformation programs? Or
  are they technical users who are unfamiliar with lens programs? I initially
  thought the tool is intended for the first type of users, but as I read
  through the paper it seems apparent that the authors intend to target the
  second group, especially with the requirement that users need to provide a
  regular expression description of the input/output data format. The
  requirement to do so seems to be the weakest point of the paper, especially
  given that existing tools (such as FlashFill or Dataplay [1]) do not have such
  requirement. If the tool is indeed intended for users who are not familiar
  with lens programming, then can the authors justify why lens is the
  appropriate programming model for data transformation? Wouldn't it be easier
  to dump the input/output data into a common format (say CSV, spreadsheet, or
  JSON), and simply use FlashFill or TriFacta [2] to learn the desired
  transformation program rather than providing the ? I understand that it might
  be difficult for tools like FlashFill to learn nested constructs (given its
  relational data model), but to me that doesn't quite justify the need for
  providing regular expressions as part of the input.

Discussed above

- Second, the evaluations seem weak.  The examples from FlashFill and are those
  that can already be synthesized using FlashFill by just providing input /
  output examples. Also, can the authors describe how complex are the regular
  expressions that users need to provide? Given that the same data format can be
  described using different semantically equivalent expressions, how do they
  affect the effectiveness of the synthesis process?

Discussed in part above.  Semantically equivalent expressions can have an impact
on the effectiveness of the synthesis process.  As we discuss in Speed of
Synthesis, in the Evaluation section, the difference in the presentation of the
regular expressions being mapped between is a stronger indicator of complexity
for this system.

- Finally, the real world contains many erroneous / dirty data. What happens if
  the user provide the incorrect regular expression or inconsistent input /
  output examples?

If there is possibly dirty data, then we expect this dirty data to not be
input to the transformer through the typing.  This would allow for the
unexpected data to fail, instead of silently succeed with unexpected results.
If this odd data is found to be an expected occurence, changes to the typing can
be made to address it, and a new lens can be synthesized.

- Can the authors describe what kind of examples were needed to
  achieve the efficient synthesis results? In light of the "Importance of
  Examples" subsection on p.11, it seems that choosing "clever" examples is one
  of the key requirement for the tool to succeed in synthesizing a valid
  program. In addition, is there some way for the tool to identify the error and
  perform iterative synthesis? Also, what happens if multiple lens program are
  found for a given input?

Clever examples do not need to be provided to the system.  The "Avg # Exmpl"
column gets its value from randomly generated examples, as we did not want our
implicit understanding of the synthesis algorithm to make our system look like
it needed fewer examples than it does when used by a non-expert.  When
multiple programs are possible, the one that requires the least significant
permutation is chosen.  The "Max # Exmpl" column corresponds to a permutation
generated which chooses a wrong permutation, if such a permutation is
permutation is possible.

### 2017-01-28 paper/rebuttal.txt

Thank you for your detailed feedback.  We will incorporate your many
useful suggestions when we revise our paper.

Several reviewers asked who benefits from this technology.   The
brief answer is that this technology is aimed at programmers.  If one
were to ask random programmers whether they know how to use regular
expressions, most would say yes.  In contrast, almost none of them
would know how to use lenses. The central idea of this paper is to
exploit something that most programmers know well (regular exprssions)
to help them write complex programs in a language that they do not
know well (lenses).  This help is particularly important because
lenses are *tricky* to write.  For example, consider the following
lens for translating between idealized bibtex and endnote author
formats:

@@

let bibtex_to_endnote_au
  : (lens in BIBTEXAUTHORINFO <=> TAGGEDAUTHORDEFNS) =
  del "author={"
    . ins "au - "
    . lens_swap (NAME . del ",") (lens_swap WSP NAME)* 
    . (del " and "
         . ins "\n au - "
         . lens_swap (NAME . del ",") (lens_swap WSP NAME)* )* 
    . del "}"

Writing this lens from scratch is difficult, and it is a relatively
simple lens! We will rewrite the introduction to clarify the goals of
our work.

A second concern stemmed from the lack of a formal treatment of our
implementation, in particular the function RigidSynth. In earlier
drafts, we had a formal description of this algorithm.  Regrettably,
we removed it to shorten the paper to fit the page limit.  We
inadvertently removed it from the extended report at the same time.
We will reinstate the full formal description in the extended report
and find ways to summarize it precisely in the final paper.

A third area of concern involved aspects of our evaluation,
specifically comparisons with other tools and measurements to evaluate
the heuristics we chose in searching for lenses.

A direct comparison with related tools is not possible because there
are no other tools that do bidirectional program synthesis.
Performance comparisons to similar tools such as FlashFill and
FlashExtract  are difficult because they cannot express the rich set
of permutations we tackle.

As for our choice of heuristics, we arrived at them when more naive
approaches failed.  For example, our initial search procedure, which
traversed arbitrary equivalences, could not synthesize anything beyond
simple microbenchmarks, even with the significant simplification of
not searching for potential compositions!  Without a priority queue,
we were able to synthesize complicated lenses when the source and
target formats were similar, but not when the formats differed
significantly. We agree that performance measurements showing the
importance of these aspects of our algorithm would improve the paper,
and we will work to add them.

In the remainder, we respond to individual concerns not already addressed.

### 2017-01-28 paper/rebuttal.txt

TARGET AUDIENCE

Several reviewers asked who benefits from this technology.  Briefly, the
technology is aimed at programmers [BCP: Do we really think the reviewers
might not have got this??  Or do we mean something more focused?].  The
central idea is to exploit something that most programmers know well
(regular expressions) to help them write significant programs in a language
that they do not know well (lenses).  This help is particularly important
because past experience has shown that lenses are often quite tricky to
write.  For example, consider the following lens for translating between
idealized bibtex and endnote author formats:

@@

  let bibtex_to_endnote_au
    : (lens in BIBTEXAUTHORINFO <=> TAGGEDAUTHORDEFNS) =
    del "author={"
      . ins "au - "
      . lens_swap (NAME . del ",") (lens_swap WSP NAME)* 
      . (del " and "
           . ins "\n au - "
           . lens_swap (NAME . del ",") (lens_swap WSP NAME)* )* 
      . del "}"

Writing this lens from scratch is nontrivial (mainly because of all the
swaps), and it is a relatively simple lens -- Augeas users regularly work
with lenses XXX orders of magnitude this size.

FORMALIZATION

A second concern stemmed from the lack of a formal treatment of our
implementation, in particular the function RigidSynth. In earlier drafts, we
actually had a formal description of this algorithm, but (in retrospect
regrettably) we removed it to get under the page limit and inadvertently
removed it from the extended report at the same time.  We will reinstate the
full formal description in the extended report and find a good way to
summarize it precisely in the final paper.

[BCP: Is there anything we can say about the formal description?  E.g., a
few words about how large or complex it is?  Any theorems we've proved about
it?]

EVALUATION

@@

A third area of concern involved aspects of our evaluation,
specifically comparisons with other tools and measurements to evaluate
the heuristics we chose in searching for lenses.

A direct comparison with related tools is not possible because there
are no other tools that do bidirectional program synthesis.
Performance comparisons to similar tools such as FlashFill and
FlashExtract  are difficult because they cannot express the rich set
of permutations we tackle.

As for our choice of heuristics, we arrived at them when more naive
approaches failed.  For example, our initial search procedure, which
traversed arbitrary equivalences, could not synthesize anything beyond
simple microbenchmarks, even with the significant simplification of
not searching for potential compositions!  Without a priority queue,
we were able to synthesize complicated lenses when the source and
target formats were similar, but not when the formats differed
significantly. We agree that performance measurements showing the
importance of these aspects of our algorithm would improve the paper,
and we will work to add them.

We will clarify all these points in the next rewrite.

===========================================================================
===========================================================================
The remainder of this document -- to be read only as needed -- responds in
detail to individual concerns not already addressed.

### 2017-01-28 paper/rebuttal.txt

TARGET AUDIENCE

Several reviewers asked who benefits from this technology.  Briefly, the
technology is aimed at programmers [BCP: Do we really think the reviewers
might not have got this??  Or do we mean something more focused?].  The
central idea is to exploit something that most programmers know well
(regular expressions) to help them write significant programs in a language
that they do not know well (lenses).  This help is particularly important
because past experience has shown that lenses are often quite tricky to
write.  For example, consider the following lens for translating between
idealized bibtex and endnote author formats:

  let bibtex_to_endnote_au
    : (lens in BIBTEXAUTHORINFO <=> TAGGEDAUTHORDEFNS) =
    del "author={"
      . ins "au - "
      . lens_swap (NAME . del ",") (lens_swap WSP NAME)* 
      . (del " and "
           . ins "\n au - "
           . lens_swap (NAME . del ",") (lens_swap WSP NAME)* )* 
      . del "}"

Writing this lens from scratch is nontrivial (mainly because of all the
swaps), and it is a relatively simple lens -- Augeas and Boomerang users
regularly work with lenses XXX orders of magnitude [BCP: quantify, maybe
with a couple examples.] this size.

@@

FORMALIZATION

A second major concern was the lack of a formal treatment of our
implementation, in particular the function RigidSynth.

In earlier drafts, we actually had a formal description of this algorithm,
but (in retrospect regrettably) we removed it to get under the page limit
and inadvertently removed it from the extended version at the same time.  We
will reinstate the full treatment in the extended report and find a good way
to summarize it in the final paper.

@@

[BCP: Is there anything we can say about the formal description?  E.g., a
few words about how large or complex it is?  Any theorems we've proved about
it?]

EVALUATION

A third area of concern was our evaluation, specifically the lack of
comparisons with other tools or a quantitative evaluation of the heuristics
we chose in searching for lenses.

Direct comparison with related tools is difficult because there aren't any
tools that do the same task: this is the first attempt at bidirectional
program synthesis.  Even performance comparisons to similar tools such as
FlashFill and FlashExtract are difficult because they cannot express the
rich set of permutations we tackle.  [BCP: This paragraph feels a bit
argumentative.  Aren't there ANY quantitative comparisons that we could do,
even if they didn't teach us too much?]

On the other hand, we agree that performance measurements quantifying the
importance of these aspects of the various heuristics used by the algorithm
would improve the paper.  It will not be hard to instrument the
implementation so that we can disable heuristics individually and gather
these numbers.  (Anecdotally, we invented each of these heuristics when more
naive approaches failed.  For example, our initial search procedure, which
traversed arbitrary equivalences, could not synthesize anything beyond
simple microbenchmarks, even with the significant simplification of not
searching for potential compositions!  Without a priority queue, we were
able to synthesize complicated lenses when the source and target formats
were similar, but not when the formats differed significantly.)

@@

===========================================================================
===========================================================================
The remainder of this document -- to be read only as needed -- responds in
detail to individual concerns not already addressed.

==========================
Review A:
==========================

- The synthesized lenses are not shown or discussed in much detail,
  nor are they compared to hand-written ones (when possible).

Our implementation puts some work into making the synthesized lenses
readable.  The transformation from DNF lenses to Boomerang lenses alone
creates fairly unintelligible syntax.  This is an issue, we would like for
programmers to be able to validate the synthesized programs through a manual
inspection of them.  As such we apply a number of optimizations, including,
but not limited to:

### 2017-01-28 paper/rebuttal.txt

TARGET AUDIENCE

Several reviewers asked who benefits from this technology.  Briefly, the
technology is aimed at programmers [BCP: Do we really think the reviewers
might not have got this??  Or do we mean something more focused?  E.g.,
Review E is not really asking who the target audience is.  Nor is review F.
Moreover, we don't really address review F's concern -- that providing
regular expressions is already onerous.  Indeed, we agree with this, and
ultimately we hope to synthesize those too; what we're reporting here is
addressing what seemed like the hardest and most novel piece of the problem;
inferring regularly expressions is a separable, and previously studied,
issue.].  The central idea is to exploit something that most programmers
know well (regular expressions) to help them write significant programs in a
language that they do not know well (lenses).  This help is particularly
important because past experience has shown that lenses are often quite
tricky to write.  For example, consider the following lens for translating
between idealized bibtex and endnote author formats:

@@

A third area of concern was our evaluation, specifically the lack of
comparisons with other tools or a quantitative evaluation of the heuristics
we chose in searching for lenses.

Direct comparison with related tools is difficult because there aren't any
tools that do the same task: this is the first attempt at bidirectional
program synthesis.  Even performance comparisons to similar tools such as
FlashFill and FlashExtract are difficult because they cannot express the
rich set of permutations we tackle.  [BCP: This paragraph feels a bit
argumentative.  Aren't there ANY quantitative comparisons that we could do,
even if they didn't teach us too much?]

On the other hand, we agree that performance measurements quantifying the
importance of these aspects of the various heuristics used by the algorithm
would improve the paper.  It will not be hard to instrument the
implementation so that we can disable heuristics individually and gather
these numbers.  (Anecdotally, we invented each of these heuristics when more
naive approaches failed.  For example, our initial search procedure, which
traversed arbitrary equivalences, could not synthesize anything beyond
simple microbenchmarks, even with the significant simplification of not
searching for potential compositions!  Without a priority queue, we were
able to synthesize complicated lenses when the source and target formats
were similar, but not when the formats differed significantly.)

MORE THINGS TO DISCUSS

[BCP: Here are some other points in the reviews that I thought were
important enough to deserve discussion in the "above the fold" part of the
response...

* the current approach works for the subset of lenses that are bijective. It
  would be good to give a sense about how much this limits its use among,
  say, existing Augeas programs. And is there reason to believe that the
  approach will or will not be amenable for larger subsets of Boomerang?

@@

Agreed, while we use similar techniques, they are used in different ways,
identifying the similarities and noting the differences would strengthen our
related work.

==========================
Review D:
==========================

- The algorithmic contributions (particularly, type-directedness) are often
  unclear.

Discussed in part above.

[BCP: Do you mean "above" as in the very top section?  Or somewhere else in
the specific review comments?  I doubt if anybody is going to read all our
responses to all the reviews, so best to make the pointers more precise --
"See response to review A..."]

RigidSynth is entirely type directed.  We use an ordering on the subcomponents
of the types to find which subcomponents must be mapped to each other.  After
these lenses are found, they are then combined into the final lens in a way
that keeps the lens well typed.

@@

We feel that requiring well aligned regular expressions is too large a burden on
the user.  The user themselves needs to think through various equivalence rules,
which we think is better suited to an automated tool.

[BCP: Give an example!]

The idea of adding in the ability to annotate fields is an interesting one.
The way that examples are embedded in the regular expressions is similar to an
annotation.  In certain scenarios, it may also be easier to give annotations
than to give a covering example.  Likely the best solution would be one which
uses both annotations and examples to help differentiate similar fields.

- Is there a restriction that the user-provided regular expressions need to
  parse the source and target formats unambiguously? For example, if a user
  provides the regular expression for a format as Name* Name*, and the input is
  John Conway, would the system disallow such user inputs?

There is such a restriction (though we allow for ambiguity when hidden under
user defined regular expressions), but the tool does not currently check that
this restriction is satisfied.

[BCP: Really???  It should, but in any case I think we don't need to mention
that it doesn't here!]

Methods for performing these checks are well studied,
and described in Boomerang as well as in Nate Foster's thesis.  An end goal for
this tool is to be integrated with the Boomerang codebase, and used as part of
the development for Boomerang programs.  Such a production quality version of
this tool would certainly provide these capabilities.

[BCP: Doesn't answer the question very clearly.]

### 2017-01-28 paper/rebuttal.txt

TARGET AUDIENCE

Several reviewers asked who benefits from this technology.  Briefly, the
technology is aimed at programmers, to be used as a code writing tool.
Previous work on data wranglers works on easily transforming data, which
can then be analyzed manually.  This work aims to provide a tool which can
assist in writing product quality data transformers.

For this work, the central idea is to exploit something that most programmers
know well (regular expressions) to help them write significant programs in a
language that they do not know well (lenses).  This help is particularly
important because past experience has shown that lenses are often quite
tricky to write.  For example, consider the following lens for translating
between idealized BibTeX and EndNote author formats:

@@

  let bibtex_to_endnote_au
    : (lens in BIBTEXAUTHORINFO <=> TAGGEDAUTHORDEFNS) =
    del "author={"
      . ins "au - "
      . lens_swap (NAME . del ",") (lens_swap WSP NAME)* 
      . (del " and "
           . ins "\n au - "
           . lens_swap (NAME . del ",") (lens_swap WSP NAME)* )* 
      . del "}"

Writing this lens from scratch is nontrivial (mainly because of all the
swaps), and it is a relatively simple lens -- Augeas and Boomerang users
regularly work with lenses significantly larger than this.  For example,
Boomerang's full version of an idealized BibTeX converter is over 300 lines of
code, and all but 4 of the studied Augeas programs were over 50 lines of code,
with 4 over 100 lines of code.

Furthermore, we agree that providing regular expressions does require the
programmers to do additional work.  While we would eventually like to find ways
to encorporate our work with existing research on inferring regular
expressions, we wanted to initially address the novel portion of this eventual
goal.  We think that even without the inference of regular expressions, this
work already is useful as a computer assisted bidirectional programming tool.

@@

FORMALIZATION

A second major concern was the lack of a formal treatment of our
implementation, in particular the function RigidSynth.

In earlier drafts, we actually had a formal description of this algorithm,
but (in retrospect regrettably) we removed it to get under the page limit
and inadvertently removed it from the extended version at the same time.  We
will reinstate the full treatment in the extended report and find a good way
to summarize it in the final paper.

The formal description is not very complex, but takes a good amount of space to
complete.  The formal description of RigidSynth first describes an ordering on
atoms, sequences, and DNF regular expressions.  Next, this ordering is used to
define a normalization procedure for DNF regular expressions.  Finally, a
procedure for "zipping" two DNF regular expressions together is provided.
If these two DNF regular expressions can be zipped, then there is a lens between
them.  A transformation from zipped DNF regular expressions and the
permutations that normalize them is then provided. These formalizations are not
very complex, but they take up a lot of space.

@@

EVALUATION

A third area of concern was our evaluation, specifically the lack of
comparisons with other tools or a quantitative evaluation of the heuristics
we chose in searching for lenses.

Direct comparison with related tools is difficult because there aren't any
tools that do the same task: this is the first attempt at bidirectional
program synthesis.  Even performance comparisons to similar tools such as
FlashFill and FlashExtract are difficult because they cannot express the
rich set of permutations we tackle.  We recognize that a quantification of this
increase in expressibility would be important for readers of our paper.  We
think that we should provide an analysis detailing how many of our programs are
inexpressible using other string synthesis tools.

Furthermore, we strongly agree that performance measurements quantifying the
importance of these aspects of the various heuristics used by the algorithm
would greatly improve the paper.  It will not be hard to instrument the
implementation so that we can disable heuristics individually and gather
these numbers.  (Anecdotally, we invented each of these heuristics when more
naive approaches failed.  For example, our initial search procedure, which
traversed arbitrary equivalences, could not synthesize anything beyond
simple microbenchmarks, even with the significant simplification of not
searching for potential compositions!  Without a priority queue, we were
able to synthesize complicated lenses when the source and target formats
were similar, but not when the formats differed significantly.)

BIJECTIVE LENS RESTRICTION

Our program is unable to synthesize general Boomerang lenses, but rather only
those which express bijections.
There certainly are limitations to restricting ourselves to bijective lenses.
Indeed, while all existing Boomerang and Augeas lenses can be transformed into
bijective lenses, few are bijective by default, often because of a lack of
bijectivity guarantees on whitespace.
We chose to restrict our domain to bijective programs as a simplification
technique for this new research, but think there are many directions this work
can be expanded to, and we feel this research is amenable to be expanded.

In particular, bijective quotient lenses, lenses which are bijective modulo an
equivalence on the source and target languages, are an expansion to this work
that already has shown promise.  Merely adding quotienting makes this tool more
practical.  XSugar - an existing, powerful language for XML and string
bidirectional transformations - expresses bijections modulo equivalences.
While we have only seen preliminary results so far, it seems that only minor
modifications were needed for our tool to generalize to bijective quotient
lenses.

@@

- The experimental evaluation is limited and does not back the claimed
  contributions.  Similar to (a), what is the effect of type-directedness on
  the experimental results?

Discussed in part in the general section.
We did not ever try an approach that was not type-directed, as we do not think
enumerative search followed by a pruning of ill typed terms would scale well.

- The fact that the avg # of examples is often 0 seems to indicate that the
  benchmarks are simple. This is perhaps because the language is very
  restricted; i.e., there are no user-defined lenses, only the core ones. Could
  you comment on that?

We discuss why the average number of examples is so low in Importance of
Examples in the evaluation section.  To highlight how complicated the input
specifications can be, the specification for the BibTeX to
EndNote conversion problem - one of our 10 least complex input formats - and the
specification for cron - our most complex format - are given
below.  We are happy to provide all our specifications.

@@

      cron_lens = [CRON_DICT <=> CRON {}]

==========================
Review E:
==========================

- The current presentation doesn’t explain what are the difficult challenges for
  this problem. Given the source and target regular expressions, and especially
  with the strongly unambiguous constraint, it seems a simple enumerative search
  should work well. I was wondering if the authors tried simpler enumerative
  solvers baselines for this problem?

Discussed in the general section.

- The second big concern is that it was not obvious who is the target audience
  for this system. If we are expecting a user to provide the regular expressions
  for source and target formats, why can’t they also write the transformation as
  well? Can the authors provide the Boomerang program for the BibTeX example and
  explain why it would be challenging for a user to write this Boomerang program
  if they can write regular expressions for the two formats.

Discussed in the general section.

- For the BibTeX to EndNote transformation example, the paper presents an
  algorithmic way to align the regular expressions to find corresponding
  permutations of sub-expressions to match different fields. Why can’t the user
  provide regular expressions in first place that are aligned? For example,
  there is an epsilon in the BibTeX regular expression, but there is no epsilon
  in the EndNote format. Wouldn’t it be a more clear and a better specification
  on user’s part to write a regular expression where EndNode format also has an
  epsilon with an Or in front. For finding permutations as well, it isn’t clear
  why can’t the user simply annotate or label different fields using some user
  interface, rather than the algorithm having to search over different matching
  of sub-expressions.

@@

We feel that requiring well aligned regular expressions is too large a burden on
the user.  The user themselves needs to think through various equivalence rules,
which we think is better suited to an automated tool.

This is very difficult when the equivalences that need to be exploited are
nonobvious or hidden away.  Consider the two regular expressions (A|B)C*
and (A|C(C*)A)|(B|B(C*)C).  It is nontrivial for a user to identify that the
equivalences that must be applied to reach two well aligned lenses.
Furthermore, the aligned regular expressions could become difficult to read and
maintain: (A|B)(e+C+CC*C) is less understandable than (A|B)C*.

@@

The idea of adding in the ability to annotate fields is an interesting one.
The way that examples are embedded in the regular expressions is similar to an
annotation.  In certain scenarios, it may also be easier to give annotations
than to give a covering example.  Likely the best solution would be one which
uses both annotations and examples to help differentiate similar fields.

- Is there a restriction that the user-provided regular expressions need to
  parse the source and target formats unambiguously? For example, if a user
  provides the regular expression for a format as Name* Name*, and the input is
  John Conway, would the system disallow such user inputs?

The regular expressions provided to the synthesis algorithm must be unambiguous.
However, we provide machinery to allow ambiguous user-defined regular
expressions to safely be subexpressions of the input format descriptors.
Using the keyword "abstract" on the definition of a user-defined regular
expression allows these abstract regular expressions to never be expanded into
their subcomponents in the search procedure.  With this opacity, ambiguous
regular expressions can safely be subexpressions of an input format descriptor,
though only the identity transformation can act on them.

### 2017-01-28 paper/rebuttal.txt

- Have you considered using a SAT/SMT solver to search the space of
  permutations?

Because of our use of orderings, we are able to find the permutations in
O(n*log(n)) time.

- The experimental evaluation is limited and does not back the claimed
  contributions.  Similar to (a), what is the effect of type-directedness on
  the experimental results?

Discussed in part in the general section. We did not ever try an approach that
was not type-directed, as we do not think enumerative search followed by a
pruning of ill typed terms would scale well, as there are many ill typed terms.

### 2017-01-29 paper/rebuttal-old.txt

TARGET AUDIENCE

Several reviewers asked who benefits from this technology.  Briefly, the
technology is aimed at programmers, as an interactive tool for creating complex
bidirection programs. Previous automated data wranglers transform and extract
data, which is then manually analyzed and used.  Unexpected data is handled
through manual inspection of the outputs, with the program potentially raising
flags when it encounters new and unexpected cases.  For some environments this
is sufficient, but this interaction model doesn't work well for published,
production quality code.  The data wrangler can't simply be resynthesized to
handle the new cases, instead the generated program must be explicit about what
types of data it can handle, and must be guaranteed to work on this data.  This
work aims to provide a tool which can assist in writing product quality data
transformers.

@@

For this work, the central idea is to exploit something that most programmers
know well (regular expressions) to help them write significant programs in a
language that they do not know well (lenses).  This help is particularly
important because past experience has shown that lenses are often quite
tricky to write.  For example, consider the following lens for translating
between idealized BibTeX and EndNote author formats:

  let bibtex_to_endnote_au
    : (lens in BIBTEXAUTHORINFO <=> TAGGEDAUTHORDEFNS) =
    del "author={"
      . ins "au - "
      . lens_swap (NAME . del ",") (lens_swap WSP NAME)* 
      . (del " and "
           . ins "\n au - "
           . lens_swap (NAME . del ",") (lens_swap WSP NAME)* )* 
      . del "}"

Writing this lens from scratch is nontrivial (mainly because of all the
swaps), and it is a relatively simple lens -- Augeas and Boomerang users
regularly work with lenses significantly larger than this.  For example,
Boomerang's full version of an idealized BibTeX converter is over 300 lines of
code, and all but 4 of the studied Augeas programs were over 50 lines of code,
with 4 over 100 lines of code.

Furthermore, we agree that inputting regular expressions and examples is a
harder task than merely inputting examples.  While we would eventually like to
find ways to encorporate our work with existing research on inferring regular
expressions, we wanted to initially address the novel portion of this eventual
goal.  We think that even without the inference of regular expressions, this
work already is useful as a computer assisted bidirectional programming tool.
